---
title: "Assignment 2: Wrapping up regression"
author: 
  - "Selena Buttery"
  - "DS303, SP25"
  - "Prof. Amber Camp"
date: 2/28/25
format: html
toc: true
editor: visual
theme: spacelab
---

# Assignment 2: Wrapping up linear and logistic regression\*

Assignment 2 covers linear and logistic regression models and includes many of the topics we have covered over the entire semester. You will be using `palmerpenguins` data, which includes a nice variety of continuous and categorical predictors.

To submit this assignment, render this file and save as a pdf. Upload the pdf to Canvas.

\*This assignment does not include mixed effects models, but you will be seeing that on another assignment.

## Load Packages

```{r}
install.packages("lme4")

```

```{r, message = FALSE}
library(lme4)
library(tidyverse)
# install.packages("palmerpenguins") # install if needed
library(palmerpenguins)
```

## Load Data

Load the `penguins` data and examine it below (use `summary()`, etc.)

```{r}
penguins <- penguins

summary(penguins)
```

## Question 1: Describe the data

What data is contained in this data set? Describe at least four variables (excluding `year`), including what they represent and their data type. Lastly, describe whether you think `year` would be a useful predictor in this data.

Data Contained in the data set:

-   Species: The species fo the penguin (Adelie, Chinstrap, Gentoo)

    -   Data Type: Factor, categorical variable

-   Island: The island where each penguin was observed (Biscoe, Dream, Torgersen)

    -   Data Type: Factor, categorical variable

-   bill_length_mm: The length of the penguin's bill measured in millimeters

    -   Data Type: Numeric

-   bill_depth_mm: the depth of the penguin's bill measured in millimeters

    -   Data Type: Numeric

-   Year: This would not be a useful predictor of the data because it does ot directly influence the physical characteristics of the penguins like bill length, bill depth, flipper length, and body mass. These traits are more likely to be influenced by the other variables like species, island location, and sex. It could potentially be a useful predictor if we are looking at trends or shifts in the populations overtime though.

## Question 2: EDA

Explore your data visually. Create at least two visualizations that show the relationship between `flipper_length_mm` and its potential predictors.

```{r, warning = FALSE}
library(ggplot2)

ggplot(data = penguins, aes(x = bill_length_mm, y = flipper_length_mm)) +
  geom_point(aes(color = species), alpha = 0.7) +
  labs(title = "Scatter Plot of Flipper Length vs Bill Length")

  theme_minimal()
```

```{r}
ggplot(data = penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot()
```

## Question 3: Apply a linear regression

Fit a simple linear regression model predicting `flipper_length_mm` from `body_mass_g`. Interpret the slope and intercept.

```{r}
linear_model <- lm(flipper_length_mm ~ body_mass_g, data = penguins)

summary(linear_model)

```

Interpret your model output in your own words below. Be sure to include a sentence explaining how `body_mass_g` impacts `flipper_length_mm` and whether or not the effect is significant.

**Answer:** Intercept: 136.7

The intercept represents that the model predicts a flipper length of 136.7 mm if the body mass of the penguin were zero. This is significant since the p value is 2e-16.

Slope = 0.01528

For each one gram of body mass, the flipper length increases by approx 0.01528 mm. This shows a positive relationship between body mass and flipper length.

## Question 4: Apply a multiple linear regression

Fit a linear regression model predicting `flipper_length_mm` from both `body_mass_g` and `bill_length_mm`. Interpret the slopes and intercept.

```{r}
multiple <- lm(flipper_length_mm ~ body_mass_g + bill_length_mm, data = penguins)

summary(multiple)

```

Similar to Question 3, interpret the model output in your own words here:

**Answer:**

Intercept = 122

When both body mass and bill length are zero, the predicted value of flipper length is 122 mm. This is statistically significant given the p value.

Slope (body_mass_g): 0.01305

For each one gram of body mass, the flipper length is predicted to increase by 0.01305 mm, this shows a positive relationship between body mass and flipper length. This value is statistically significant according to the p value.

Slope (bill_length_mm) = 0.5492

This slope shows that for each additional millimeter in bill length, the flipper length is predicted to increase by 0.5492 mm. This shows a positive relationship and the p value shows that this value is statistically significant.

Question 5: Include an interaction

Fit a simple linear regression model predicting `flipper_length_mm` from `body_mass_g,` `bill_length_mm`, and the interaction of the two. Interpret the slopes and intercept.

```{r}

interaction <- lm(formula = flipper_length_mm ~ body_mass_g * bill_length_mm, data = penguins)

summary(interaction)
```

Interpret the model output in your own words below. If there was a change in the pattern of significance, try to explain the logic below as well.

**Answer:**

Intercept: 170.6 mm

When both body mass and bill length are zero, the predicted flipper length is 170.6 mm. The p value shows that this is statistically significant.

Slope for Body Mass: 0.00043

This is showing that for each one gram increase in body mass, the flipper length goes up 0.00043 mm. But this is not statistically significant because the p value is 0.916.

Slope for Bill Length: -0.5051

This is showing that for each one mm increase in bill length, the flipper length decreases by .5051, but this is not statistically significant since the p-value is .153.

Interaction (body mass and bill length): 0.00027

This shows that when both body mass and bill length increase together , flipper length is predicted to increase as well by 0.00027. This is statistically significant because the p-value is 0.00234.

There was a change in the pattern of significance between the previous model and this one because previously both body mass and bill length were individual;y good predictors of flipper length, but in this model their individual effects on flipper length were not significant, but the interaction was. This suggests that the relationship between flipper length and body mass/bill length are not additive, body mass and bill length influence flipper length in a way that is not captured by their individual effects.

## Question 6: Compare models

Compare the models you built in Questions 4 and 5 using `anova()`.

```{r}
anova(multiple, interaction)

```

-   Which is the better model? How do you know?

    -   **Answer:** The interaction model is better because it is being shown as significant according to the p-value which is 0.002341.

-   Is it possible to compare the models from Questions 3 and 5 using the same method? Why or why not?

    -   **Answer:** No, it is not possible to compare the models from Question 3 and 5 using ANOVA because question 3 model is a simple linear regression while question 5 includes two predictors an an interaction term.

## Question 7: Categorical predictors

Build a linear model that includes a categorical predictor of your choice. It is fine to stick with dummy coding. Optional: apply a different coding scheme AND interpret the output correctly for +1 extra credit.

```{r}
  
penguins$species_helmert <- as.factor(penguins$species)
contrasts(penguins$species_helmert) <- contr.helmert(3)


lm_helmert <- lm(flipper_length_mm ~ species_helmert, data = penguins)

summary(lm_helmert)

```

-   What is the reference level of your categorical predictor?

    -   **Answer: The Adelie species**

-   What is your interpretation of this model output? Address all coefficients.

    **Answer:**

-   Intercept: 200.99

    -   This intercepts that on average, Adelie penguins have a flipper length of 200.99 mm.

-   species_helmert1: 2.9349

    -   This coefficient represents the difference in flipper length between Chinstrap and Adelie penguins, the coefficient is positive so this indicates that Chinstrap penguins have a flipper length that is predicted to be 2.93 mm longer than Adelie penguins. The p-value (3.79e-09) shows that this coefficient is statistically significant.

-   species_helmert2: 8.0995

    -   Similar to the one above, this represents the difference in flipper length between Gentoo and Adelie penguins. The coefficient is positive so this indicates that Gentoo penguins have a flipper length around 8.10 mm longer than Adelie penguins and this is statistically significant according to the p-value of 2e-16.

## Question 8: Relevel your categorical variable

Relevel your categorical variable so that a **different** level becomes the reference. Then, run the same model you did in Question 7 and interpret the output.

Relevel:

```{r}

penguins$species <- relevel(penguins$species, ref = "Chinstrap")
```

Apply model from Question 7:

```{r}

lm_relevel <- lm(formula = flipper_length_mm ~ species, data = penguins)

summary(lm_relevel)

```

-   What is the new reference level of your categorical predictor?

    -   **Answer: Chinstrap**

-   What is your interpretation of this new model output? Address all coefficients.

    **Answer:**

-   Intercept: 195.8235

    -   This is the average flipper length in mm of Chinstrap penguins.

-   speciesAdelie: -5.8599

    -   This coefficient shows that adelie penguins have a 5.87 mm shorter flipper length on average than Chinstrap penguins. This is also statistically significant according to the p-value.

-   speciesGentoo: 21.3635

    -   This coefficient shows that Gentoo penguins have a 21.36 mm longer flipper length on average than Chinstrap penguins and this is also statistically significant according to the p-value.

## Question 9: Apply a logistic regression

Apply a logistic regression. Include as many predictor variables as you'd like. Remember that your predicted outcome variable needs to be binary (or categorical with two levels).

\
Hint: You could use `sex` or create a binary variable of your own (e.g., Gentoo vs. non-Gentoo) to test your model.

```{r}

penguins_subset <- subset(penguins, species %in% c ("Adelie", "Chinstrap"))

penguins_subset$species_num <- ifelse(penguins_subset$species == "Adelie", 0, 1)

penguins_model <- glm(species_num ~ bill_length_mm + flipper_length_mm, data = penguins_subset, family = binomial)

summary(penguins_model)
```

What are your key takeaways from this model?

**Answer:**

-   Bill length is the most important factor in predicting if a penguin is a Gentoo.

-   Flipper length has a negative effect, slightly decreasing the chance of being a Gentoo.

-   The model fits well, it helps to explain the variance in species classification.

Question 10: Synthesize the information

Imagine you’re a biologist studying penguin populations. Which predictors do you think are most important to measure or record in the field to predict flipper length? Why?

**Answer:** I would say that the most important predictors to measure in the field to predict flipper length would be body mass, bill length, and species. Body mass because heavier penguins tend to have longer flippers which would make this a strong predictor. There is a measurable relationship between bill size and flipper length probably due to the different species. And lastly species because different penguins species have distinct body proportions which would make species identification crucial for more accurate preictions.

## Bonus: Stepwise Regression

Perform stepwise regression to find the best model for an outcome of your choice. You will likely encounter an error – fixing that error and explaining your findings will earn you +1 extra credit. Show your work.

```{r}

model <- lm(flipper_length_mm ~ bill_length_mm + bill_depth_mm + body_mass_g + species + sex, data = penguins)
```

```{r}
stepwise <- stepAIC(model, direction = "both", trace = TRUE)
```

Fixing error:

```{r}
penguins_clean <- na.omit(penguins)
```

```{r}
model <- lm(flipper_length_mm ~ bill_length_mm + bill_depth_mm + body_mass_g + species + sex, data = penguins_clean)
```

```{r}
best_model <- stepAIC(model, direction = "both", trace = TRUE)
```

According to this stepwise regression, explain how the final model was selected.

**Answer:** According to the stepwise regression, the initial model included bill length, bill depth, body mass, species, and sex as predictors. Stepwise regression removed sex first because it was not a significant predictor. After removing sex, the model included bill length, bill depth, body mass, and species and had an AIC of 1107.7. Removing any more of these variables increased the AIC which makes the model worse.
